\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-08-22 phd-cs-us.tex
Updates from Overleaf

\title{\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A guide for International Students}}
%\date{\today}
\author{\small \href{https://nguyenthanhvuh.github.io}{ThanhVu (Vu) Nguyen} (\href{mailto:tvn@gmu.edu}{tvn@gmu.edu})}

@@

@@

\tcblower
This shows the student has the persistence and determination required for research. I would also advocate for accepting that student!
\end{commentbox}

@@

\subsection{How to rank or select a CS graduate program?}
   \begin{center}
    \includegraphics[scale=0.4]{c1.png}
   \end{center}

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school that you didn't know much about can have very strong research in your interested topic (and vice versa, a school you thought highly about has no one works in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and probablyo nly in CS), conferences, not journals, are often the main venue to publish research finding.}. \S\ref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

@@

\begin{wrapfigure}{r}{0.48\textwidth}
\vspace{-0.2in}
      \begin{center}
        \includegraphics[width=0.48\textwidth]{c8.png}
      \end{center}
      \vspace{-0.2in}
    \end{wrapfigure}
The short answer is that tenured-track faculty such as assistant professors are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding (yet).

Tenured faculty, e.g., associate and full profs., are more likel older, more well-known, and have more experience in managing students.  However, they might not push you as hard (they don't have to, they already got tenured). They might also expect you to figure things out yourself (so you need to be very independent).  Some tenured faculty are also no longer active in research and more involved with administrative duties or with their startup companies.

@@

Absolutely, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able to leverage this to make your profile stand out as mentioned in \S\ref{sec:improve-your-chance}.

@@

\begin{wrapfigure}{r}{0.28\textwidth}
\vspace{-0.2in}
\begin{center}
\includegraphics[width=0.28\textwidth]{c4a.png}
\end{center}
\vspace{-0.1in}
\end{wrapfigure}
    Typically, 5--7 years for PhD in CS in the US.  This can be longer than CS PhD at universities in other countries, which might require MS first (recall that US PhD programs do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students in the US also often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

    The first 2 years you spend on coursework, finding an adviser, learning research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job. 
    In many cases you might take a summer or two off to do internship to get additional research opportunities.
    The PhDComics figure on the right shows the ``ambition'' level of a PhD student over their years of study (they miss the 6-7th Year where the ambition is \emph{``Just let me graduate''}).

\subsection{PhD in other related field such as CE, IST, Cybersecurity, and Stats}

In many cases you do not need to do a PhD in CS to work in your area of interested. For example, in addition to having a traditional CS department, GMU has IST and Cybersecurity departments, which have faculty  with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  In short, it is totally possible that you still get to do CS research and publish in CS-related venues even if you're not in a traditional CS program.  It is very common to see faculty with CS PhD in a non-CS department as well as faculty with non-CS PhD in CS department.

@@

\begin{wrapfigure}{l}{0.5\linewidth}
\vspace{-0.3in}
      \begin{center}
        \includegraphics[width=1.0\linewidth]{c5.png}
      \end{center}
      \vspace{-0.5in}
    \end{wrapfigure}
If you don't know the professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste names, so no need to do so,  just Prof. or Dr. Lastname.

>>>>

\title{\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students}}
%\date{\today}
\author{\small \href{https://nguyenthanhvuh.github.io}{ThanhVu (Vu) Nguyen} (\href{mailto:tvn@gmu.edu}{tvn@gmu.edu})}

@@

For Vietnamese students, it's worth highlighting the \href{https://vef2.org/}{VEF 2.0 program}, which has played a crucial role in aiding many good students in gaining acceptance to top-tier PhD programs in the US. VEF 2.0 follows an intriguing model where  US faculty members from leading institutions are invited to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily confined to the interview itself), the impact of these reference letters is remarkably positive. This fact is underscored by the noteworthy success of numerous VEF 2.0 students in securing admission to top CS PhD programs across the US.

@@

\tcblower
I would also advocate for that student as this shows they have the persistence and determination required for research.
\end{commentbox}

@@

\subsection{How to rank or select a CS graduate program?}
   \begin{center}
    \includegraphics[scale=0.5]{c1.png}
   \end{center}

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school that you didn't know much about can have very strong research in your interested topic (and vice versa, a school you thought highly about has no one works in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research finding.}. \S\ref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

\begin{commentbox}[Hung]
I always encourage the students I admitted to talk with my students and the students of other faculty in other schools who admitted them. You will unlikely hear straight-out complaints from current students in a professor’s group. But sometimes what is important are things that they (current students) don’t tell you. Pay attention to their "level of excitement" being in the group.
\end{commentbox}

@@

\begin{wrapfigure}{r}{0.48\textwidth}
\vspace{-0.2in}
      \begin{center}
        \includegraphics[width=1\linewidth]{c8.png}
      \end{center}
      \vspace{-0.2in}
    \end{wrapfigure}
The short answer is that tenured-track faculty such as assistant professors are more likely to be young and active in research (they have to, in order to get tenure). Thus, they will likely have more time to work with you and push you to do research and publish. However, they may not have as much experience in managing students and may not have as much funding (yet).

Tenured faculty, e.g., associate and full profs., are likely older, more well-known, and have more experience in managing students.  However, they might not push you as hard (they don't have to, they already got tenured). They might also expect you to figure things out yourself (so you need to be very independent).  Some tenured faculty are also no longer active in research and more involved with administrative duties or with their startup companies.

@@

Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able to leverage this to make your profile stand out as mentioned in \S\ref{sec:improve-your-chance}.

@@

\begin{wrapfigure}{r}{0.35\textwidth}
\vspace{-0.2in}
\begin{center}
\includegraphics[width=1\linewidth]{c4a.png}
\end{center}
\vspace{-0.3in}
\end{wrapfigure}
    Typically, 5--7 years for PhD in CS in the US.  This can be longer than CS PhD at universities in other countries, which might require MS first (recall that US PhD programs do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students in the US also often take a ``leave of absence'' for 1--2 semesters to do internship at companies and research labs.

    The first 2 years you spend on coursework (somewhat equivalent to an MS study), finding an adviser, learning research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job. 
    In many cases you might take a summer or two off to do internship to get additional research opportunities.
    The PhDComics figure on the right shows the ``ambition'' level of a PhD student over their years of study (they miss the 6-7th Year where the ambition is \emph{``Just let me graduate''}).

\subsection{PhD in other related field such as CE, IST, Cybersecurity}

In many cases you do not need to do a PhD in CS to work in your area of interested. For example, in addition to having a traditional CS department, GMU has IST and Cybersecurity departments, which have faculty  with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So it is totally possible that you still get to do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is very common to see faculty with CS PhD in a non-CS department as well as faculty with non-CS PhD in CS department.

@@

\begin{wrapfigure}{l}{0.5\linewidth}
\vspace{-0.3in}
      \begin{center}
        \includegraphics[width=1\linewidth]{c5.png}
      \end{center}
      \vspace{-0.5in}
    \end{wrapfigure}
If you don't know the professor (e.g., first email contact), then use \textbf{Prof. Lastname} or \textbf{Dr. Lastname}. I've seen many international students write \textbf{Prof.} or \textbf{Dr.} \textbf{Firstname Lastname}.  Writing like that makes it like you copy and paste names, so no need to do so,  just Prof. or Dr. Lastname.
